\chapter{Presentation of the novel \textit{Marta}}
This chapter describes the novel \textit{Marta} by Eliza Orzeszkowa.
It contains a brief presentation of the novel's historical background: the Polish people's struggle for independence, the literary trend called ``Warsaw Positivism,'' and the life story of the author.
Finally, the plot of the novel is outlined.

\section{Historical background of the novel}

In order to fully understand the works of Eliza Orzeszkowa, it is important to put her writings in their historical context.
Towards the end of the 18th century, the Polish-Lithuanian Commonwealth was in decline.
Taking advantage of the weaknesses of the Polish state and military, three neighboring countries gradually annexed the territory of the Commonwealth.
Three partitions of Poland have taken place, in 1772, 1793, and 1795, after which the Polish crown lost all territory to the Russian Empire, the Kingdom of Prussia, and the Austro-Hungarian Empire. Poland ceased to exist as a country.
After the partitions, the ruling forces exercised policies aiming to uproot any expression of Polish patriotism and nationalism.
These policies have been particularly strict in the Russian and Prussian partitions.

Unfortunately for the occupants, the Polish people were not willing to give up their patriotic spirit.
For 123 years, between 1795 and 1918, their struggle for independence constituted an important topic in Polish-language literature.
After the failure of the 1863 January Uprising against the Russian Empire, the Polish people were disappointed with Romanticism and slogans of armed fight for independence
% The fallen uprising and the Polish question had also been an important topic for European socialists
(Stekloff 1928).

The end of the January Uprising is considered to be the beginning of a literary and philosophical genre known as ``Warsaw Positivism'' (French: \textit{positivisme varsovien}, Polish: \textit{pozytywizm warszawski}).
This philosophy emphasized the importance of the so-called ``organic work'' (Polish: \textit{praca organiczna}), that is, active development of education and economy.
% It was particularly important in the works of Positivist writers, who believed that the key to independence was education and hard work, rather than uprisings and revolutions.
The key topics in Polish Positivist literature included the fight for independence, the emancipation of serfs and women, promoting science, medicine and public hygiene, and the assimilation of the Jewish minority with the Polish society.
Besides Eliza Oreszkowa, important representants of Warsaw Positivism included Bolesław Prus (1847--1912), Maria Konopnicka (1842--1910), and Henryk Sienkiewicz (1846--1916), the laureate of the Nobel Prize in Literature 1905
(eSzkola.pl).

Gloger (2007) states that in the case of Poland, Positivism played a similar role to that of Enlightenment in Western Europe, that is to say, it paved ground for the development of modernism and modernity, popularizing rationalism and a scientific approach to reality. On the other hand, the cultural impact of Enlightenment in Poland had been rather limited due to the overall civilizational lag and unfavorable historical circumstances.

\section{The life and times of Eliza Orzeszkowa}

Eliza Orzeszkowa was born Eliza Pawłowska on June 6th, 1841, in the village of Milkowszczyzna, north of the Niemen river, in present-day Belarus, to a family of gentry.
Her father, Benedykt Pawłowski, died when she was three years old. 
At the age of ten, she moved to Warsaw, the present-day capital of Poland, where for the following five years she attended a boarding school, run by the nuns of the Order of the Holy Sacrament.
During that period, she had studied French, German, and Polish literature.

In 1858, at the age of 17, her parents arranged her marriage with a wealthy landowner by the name of Piotr Orzeszko.
In her diaries, Eliza admits her fondness for the tall, blond, and handsome man, and was excited to leave her family home to explore the country and to visit the man's numerous relatives, living in different parts of the Russian empire.
At the day of their wedding, the bridegroom was 35 years old
(Shastouski).

The marriage of Mr. and Mrs. Orzeszko proved unsuccessful, mainly due to Eliza's political interests and affiliations---Eliza was highly pro-independence and sought the emancipation of serfs.
Her husband, on the other hand, as a member of the bourgeoisie, was more inclined to protect the interests of his class, which saw the political turmoil brought about by military activities as an opportunity to further accumulate wealth
(Britannica, Bachórz, Brykowisko, 2011, Orzeszkowa, 2000, p. 7).

% In 1864, serfdom was effectively abolished in the Russian partition --- peasants were allowed to own their land.

During the January Uprising, Eliza was actively working for the Polish cause, passing messages between the troops, and even helping Mr. Romuald Traugutt, the leader of the insurrection from October 1863 up to its end in August 1864, by hiding him in her house and escorting him to the border of Congress Poland.
Her husband, on the other hand, had been critical of the uprising from its very beginning.
He saw the uprising as untimely and highly unlikely to be successful.
He would say that the movement would only bring about unnecessary death, suffering, and grave repercussions for the nation
(Shastouski, Brykowisko, 2011).

Following the failed uprising, in December 1864, Mr. Orzeszko was arrested and sent to Russia, his land and possessions confiscated.
Deprived of her husband's estate, Eliza had no other choice than to return to her father's house in Milkowszczyzna.
Eliza has thus described her impressions from returning to her hometown:

\begin{displayquote}
  I went out to the world in a carriage padded with blue damask and drawn by seven horses. I was accompanied by servants shining with coat of arms plaques\ldots{} At the age of twenty, I returned from those distant places all alone, to the house I have inherited from my father, only to find it looted by soldiers who had stayed in the household; [I had to pay] debts taken by other people and a tribute % kontrybucja
  so enormous that even [our] fields of wheat would bend under its burden; various officials would frequently visit and overstay\ldots{} The situation was completely overwhelming\ldots{}%
\footnote{Shastouski, Orzeszkowa, 2000, pp. 7--8.}
\end{displayquote}

Eliza was not the only one to suffer in the new reality.
Following the failure of the January Uprising and the abolishment of serfdom in the Russian Empire, many families in Lithuania%
\footnote{The name ``Lithuania'' at the time of Orzeszkowa was commonly used to refer to the territories of the former Great Duchy of Lithuania, encompassing a much greater area than today's Republic of Lithuania, and including parts of present-day Belarus and West Ukraine. These regions still remain in the Polish sphere of influence and have a substantial Polish minority, although the influence was diminished by decades of Soviet rule. Even so, Poles born before WW2 still refer to the cities of Lviv, Vilnius, and Hrodna as ``old Polish cities.''}
have become impoverished.
The difficult situation gave Eliza the opportunity to watch the world from the point of view of the weak and destitute, which would prove crucial for her literary career.
In 1866, the young writer published her first short story under the title \textit{A Picture from the Years of Famine}%
\footnote{Polish: \textit{Obrazek z lat głodowych}. The English title used in this thesis is the author's own translation.}, dealing with the daily struggles of the lower layers of society.
The story was an early example of Orzeszkowa's realism, depicting impoverished peasants as innocent victims of a vicious, rotten feudal system.
On the positive side, the story called for social progress for the benefit of the society as a whole, rather than just specific social classes or castes, a direction highly unusual for Polish literature in the period directly following the failed uprising
(Shastouski, Orzeszkowa, 2000, pp. 8--11).

In other realist works, such as \textit{The Pompalinski Family} (Polish: \textit{Pompali\'nscy}), \textit{Mr. Graba} (Polish: \textit{Pan Graba}), or \textit{On the Bottom of Conscience} (Polish: \textit{Na dnie sumienia}%
\footnote{All three titles translated from Polish by the author.}), Orzeszkowa makes a point of describing and ridiculing the most annoying aspects of the elites, depicting aristocrats as a parasitic, dull-witted folk.
In \textit{Mr. Graba} and \textit{On the Bottom of Conscience}, she tells the stories of young people who have received traditional, aristocratic education, which proved useless once the material basis for their existence had disappeared---the protagonists had no other choice than to live a parasitic, opportunistic lifestyle
(Orzeszkowa, 2000, p. 11).

During a period of roughly 10 years following her creative debut, Eliza's writing had been influenced by the emerging trend of Warsaw Positivism, full of optimism for the future of a society undergoing a capitalist transformation.
The pivot of her creative works of this period had been educating the youth to adapt to the new bourgeois reality 
(Orzeszkowa, 2000, pp. 9).

The optimistic outlooks became somewhat dimmed when she realized that she herself, as a single woman, might struggle to make a livelihood after losing the main sources of her income, i. e. the estates of her husband and father, and would need to resort to the only jobs available to women, such as private language tutor or governess.
At the time, the most attractive of all jobs deemed fit for women had been the job of a telegraphist.
Even so, it soon became obvious that in Warsaw, under Russian rule, even that position was out of reach of a Polish woman
(Orzeszkowa, 2000, p. 9).

Eliza considered a woman's lack of control over her own existence to be yet another remnant of the bygone feudal era.
She had experienced this feeling of impotence herself, when she tried to apply to the Church for a divorce.
The Catholic Church, with its conservative views, was an important bastion of the feudal system, and used its arsenal of social manipulation techniques to protect the feudal privilege.
For a long time, the ecclesiastic authorities were unwilling to grant a divorce request of a woman, and only granted her wish after she agreed to pay the Church a considerable amount of money.
Following her divorce, she had been ostracized and ridiculed by the conservative members of the society.
This bitter realization of a woman's impotence is emphasized in her works of \textit{Tendenzliteratur}, including the novel \textit{Marta}, analyzed in this thesis
(Orzeszkowa, 2000, pp. 8--11).

In 1869, Eliza's marriage was finally annulled.
After selling her father's estate in Milkowszczyzna in 1970, she settled in Grodno (present-day Hrodna, Belarus).
Since 1879, she co-owned a publishing house and bookstore in Vilnius, in present-day Lithuania.
Her most important works included \textit{The Boor} (Polish: \textit{Cham}), \textit{The Dziurdzia Family} (Polish: \textit{Dziurdziowie}), and \textit{On the Banks of the Niemen} (Polish: \textit{Nad Niemnem}).
Most of her works deal with the difficult existence of peasants, women, and impoverished gentry.
Her works are often classified as \textit{Tendenzliteratur} (German: `tendentious literature').
She had been nominated twice for the Nobel Prize in Literature.
Eliza Orzeszkowa died in Grodno in 1910 and was buried at the local Catholic cemetery
(Brykowisko, 2011, Britannica, Orzeszkowa, 2000, p. 10).

\section{Names of Eliza Orzeszkowa}
In many Slavic languages, including Polish, many family names have traditionally taken a different form when referring to a man, his unmarried daughter, and his wife.
For instance, the wife of the well-known Polish poet of the Romanticist period, Adam Mickiewicz, was referred to as Celina Mickiewiczowa, and his eldest daughter would be referred to as Maria Mickiewiczówna, up until her marriage.
Analogously, in case of the name ``Eliza Orzeszkowa,'' Orzeszkowa means `the wife of Mr. Orzeszko,' which roughly corresponds to the English form ``Mrs. Piotr Orzeszko.''
In modern-day Polish, however, this convention is virtually obsolete.
All family members use the same form of the name, and the wife of \textit{Pan Orzeszko} (Mr. Orzeszko) can be called \textit{Pani Orzeszko} (Mrs. Orzeszko).
A notable exception to this rule are the surnames ending in \textit{-ski} or similar suffixes, such as Kowalski, Górecki, Grodzki.
These surnames have evolved from adjectives and still follow all the grammar rules pertaining to adjectives.
Thus, the wife and daughter of the Polish counterpart of John Smith, Jan Kowalski, would still use the form \textit{Kowalska}. 

In Zamenhof's translation of \textit{Marta}, the author's name is listed as Eliza Orzeszko.
Zhong Xianmin's translation follows this convention, rendering the name into Chinese with slightly distorted pronunciation as 愛麗莎・奧西斯哥 (\nazwisko{Àilìshā Àoxīsīgē}).
Other Chinese-language writings and websites use other variations, based on the traditional form ``Eliza Orzeszkowa.''
These names include 艾麗查・奧熱什科娃 (\nazwisko{Àilìchá Àorèshíkēwā}) and 艾麗查・奧若什科娃 (\nazwisko{Àilìchá Àoruòshíkēwā}). In the remaining part of this thesis, the form ``Eliza Orzeszkowa'' shall be used, being the most prevalent in Polish-language writings.

\section{Plot outline of the novel \textit{Marta}}
The version of the novel \textit{Marta} available at Wikisources is based on a 1907 edition published in Lwów (nowadays Lviv, Ukraine) and Złoczów (nowadays Zolochiv, Ukraine) by the Bookstore of Wilhelm Zukerkandl (Polish: \textit{Księgarnia Wilhelma Zukerkandla}).
This edition is divided into an introduction and nine chapters numbered with Roman numerals (from I to IX).
This numeration is not present in the Esperanto translation, which is only available as a continuous text.

The novel \textit{Marta} tells the tragical story of a certain Marta Świcka.
The story is told from the point of view of a third-person omniscient narrator.
In the beginning of the novel, Marta is presented as an impoverished, twenty-odd-year-old lady whose affluent husband had just died, leaving her with a little daughter, no living family members, and no means of livelihood.
Due to these unfavorable circumstances, she is compelled to move out of a lavish apartment in Warsaw to a plain, dilapidated, single room.

Throughout the novel, the protagonist struggles to find a job to provide for herself and her little daughter.
At first, she tries her luck as a teacher of French and piano, but soon finds out that the little qualifications she has are not sufficient for her to make ends meet.
She also finds out that she cannot work in certain professions despite having the necessary qualifications, because those jobs are only available to men.
She manages to get a job as a seamstress, which she soon quits due to a conflict at workplace.
Her little daughter falls ill with bronchitis.
Marta is compelled to return the rudimentary furniture that she rented for her tiny apartment.
Finally, a horse-drawn omnibus runs over Marta and she dies on the spot.

\section{Note on currency in Russian Poland}

At the time when the novel \textit{Marta} takes place, Warsaw was located in a political entity officially named ,,Kingdom of Poland'' (Russian: Царство Польское% add transcription
, Polish: Królestwo Polskie).
This polity was created by the 1815 Congress of Vienna and controlled by the Russian Empire.
It is ofter referred to as ,,Congress Poland'' (Polish: \textit{Królestwo Kongresowe}, lit. `Congress Kindgom').
The monetary system used in Congress Poland was based on gold and silver, and before the January Uprising, rouble coins minted in the Kingdom of Poland used bilingual denominations, with the traditional Polish currency name, \textit{zloty} (Polish: złoty, `golden') equivalent to 15 Russian kopeks (Russian: копейка, Polish: \textit{kopiejka}), i.\ e.\ $15/100$ of a Russian rouble.
After the January Uprising, Polish denominations were removed from the coins, but the conversion rate was still in common usage among the local populace.
Therefore, both currency names are mentioned throughout the source text.
